Performance Evaluation of Big Data Processing Strategies for Neuroimaging

### 2018-12-03 paper.tex

Big Data processing engines specialize in minimizing data transfers, through 
use of concepts such as data locality, in-memory computing and lazy evaluation,
to speed up the processing time of such large datasets.
Data locality was popularized by the 
MapReduce~\cite{dean2008mapreduce} framework. Rather than performing costly 
data transfers across the network, tasks were scheduled to the nodes where data
was located. Apache Spark~\cite{zaharia2016apache} succeeded MapReduce with its
incorporation of in-memory computing and lazy-evaluation which brought 
significant speed-ups in Big Data applications. In-memory computing
enabled data to be cached in memory whenever possible, reducing more costly
disk I/Os between pipeline steps. Introducing lazy
evaluation meant that results were only
computed only when necessary, enabling further performace optimizations. 
Frameworks such as
MapReduce and Spark have since become mainstream tools for data analytics, 
although many others, such as language-specific 
Dask~\cite{rocklin2015dask}, are emerging. 
Meanwhile, several scientific 
domains including bioinformatics, physics or astronomy, have entered 
the Big Data era due to increasing data volumes and variety. 
However, the adoption of Big Data engines for scientific data analysis 
remains limited, perhaps due to the widespread availability of 
scientific processing engines such as Pegasus~\cite{deelman2005pegasus}, 
Taverna~\cite{oinn2004taverna} and the necessary adaptations that would need to
be made to Big Data processing engines for scientific computing.

@@

\newpage % Remove this when abstract is written

@@

An RDD is an immutable parallel data structure that achieve fault-tolerance 
through the concept of lineage~\cite{RDDpaper}. Rather than permitting
fine-grained transformations, only coarse-grained transformations (applying to
many elements in the RDD), can be applied, thereby making it simple to keep a 
log of how the data was modified. This log, known as the lineage, is used
to reproduce any lost data.

Two types of operations can be performed on RDDs:
transformations and actions. Applying a transformation to an RDD produces a new,
child, RDD through a narrow or wide dependency. A narrow dependency signifies 
that the child is only dependent on a single parent partition, whereas a child 
RDD is dependent on all parent partitions in a wide dependency. Examples of 
spark transformations include map, filter and join. To materialize an RDD, an
action must be performed, such as a reduce or a collect.

All actions and wide dependencies require a shuffle -- Spark's most costly
operation. Every shuffle begins with each map task saving its data to local
files to be fetched by the shuffle operation. The shuffle operation then 
redistributes the data across the partitions as requested. A shuffle marks a 
stage boundary in Spark. In other words, a shuffle will not begin until all 
dependent map tasks have completed.

Spark will spill RDD data to disk whenever the complete dataset does not fit in
memory. Moreover, as Spark transformations generate new RDDs and numerous 
transformations may occur in a single application, Spark implements a LRU eviction
policy. If an evicted RDD needs to be reused, Spark will recomputed it using the
lineage data collected. It therefore becomes of great importance to persist to
memory and/or 
disk RDDs which will be reused throughout the application, particularly if the
recomputation is costly. 
%mention results in thunder paper

Data locality in Spark is achieved through the  scheduling task to partitions which have the
data loaded in memory. If the data is instead stored on HDFS, the scheduler will
assign it to one of the preferred locations specified by HDFS. Spark's scheduler
utilizes delay scheduling to ensure optimal fairness and locality for all tasks.
%improve paragraph

Three different types of schedulers are compatible with Spark: 1) Spark 
Standalone, 2) YARN~\cite{yarn} and 3) Mesos~\cite{mesos}. The Spark Standalone
scheduler is the default scheduler. YARN was a scheduler designed for Hadoop~\cite{hadoop}
clusters. It is prepackaged with Hadoop installations. In contrast to YARN, 
Mesos was designed to be used in multitenant cluster environments. In our experiments,
we focus on the Spark Standalone cluster.

Executing Spark applications on HPC systems with Spark-unaware schedulers may
be inefficient. The amount of resources requested by Spark may impede Spark-cluster
scheduling time. Using pilot-scheduling strategies to add nodes to the Spark cluster
as they are allocated by the underlying schedulers may speedup allocation and 
overall processing time. This, however, is not studied in the current paper.

@@

Nipype is a popular Python neuroimaging processing engine. It aims at being a 
solution for easily creating reproducible neuroimaging workflows. Although Nipype
does not employ any Big Data processing strategies, it provides access to numerous
plugins to schedulers found in most clusters readily available to researchers, 
such as SGE, PBS, SLURM and HTCondor. It also includes its own scheduler, MultiProc,
for parallel processing on single nodes. Nipype also includes many built-in 
interfaces to commonly used neuroimaging tools to be incorporated within the 
workflows. Nipype's ability to easily parallelize workflows in researcher-available 
cluster setups, capture detailed provenance information necessary for reproducibility
and allow users to easily integrate existing neuroimaging tools, make it preferable
to existing Big Data solutions, which would necessitate modification to achieve this.

Jobs, or Interfaces, in Nipype, are encapsulated by a Node. A Node dictates that
the job will execute on a single input. However, there are exists the MapNode, 
a child variant of the Node, which can execute on multiple inputs. All tasks in
Nipype execute in their own uniquely named subdirectory which enables the provenance 
tracking of inputs and outputs. This also enables checkpointing of the workflow.
In the case of Node failure or application modification, only the nodes who have
been modified (verified by hash) are re-executed.

In order for Nipype to operate as intended, a filesystem shared by all the nodes 
is required. However, it is still possible to save to a non-shared local filesystem, 
but it may come at the expense of fault-tolerance. Moreover, the user will need to
ensure that the files are appropriately directed to the nodes that require them as
there is no guarantee of data locality in Nipype.

>>>>

Big Data processing engines reduce data-related overheads, through 
use of concepts such as data locality, in-memory computing, and lazy evaluation.
Data locality-aware scheduling, popularized by the MapReduce~\cite{dean2008mapreduce} 
framework, runs tasks to the nodes holding the data rather than 
performing costly data transfers across the network. In-memory 
computing, adopted by MapReduce's successor Apache 
Spark~\cite{zaharia2016apache}, caches data in memory whenever 
possible, reducing costly disk I/Os between pipeline steps. 
Lazy evaluation, also used in Spark, means that results are computed only 
when necessary, enabling further performance optimizations. Frameworks 
such as
MapReduce and Spark have become mainstream tools for data analytics, 
although many others, such as 
Dask~\cite{rocklin2015dask}, are emerging. 
Meanwhile, several scientific 
domains including bioinformatics, physics or astronomy, have entered 
the Big Data era due to increasing data volumes and variety. 
However, the adoption of Big Data engines for scientific data analysis 
remains limited, perhaps due to the widespread availability of 
scientific processing engines such as Pegasus~\cite{deelman2005pegasus} or
Taverna~\cite{oinn2004taverna}, and the adaptations required in Big 
Data processing engines for scientific computing.

@@

@@

An RDD is an immutable parallel data structure that achieve fault-tolerance 
through the concept of lineage~\cite{zaharia2010spark}. Rather than permitting
fine-grained transformations, only coarse-grained transformations (applying to
many elements in the RDD), can be applied, thereby making it simple to keep a 
log of how the data was modified. This log, known as the lineage, is used
to reproduce any lost data.

Two types of operations can be performed on RDDs:
transformations and actions. Applying a transformation to an RDD produces a new,
child, RDD through a narrow or wide dependency. A narrow dependency signifies 
that the child is only dependent on a single parent partition, whereas a child 
RDD is dependent on all parent partitions in a wide dependency. Examples of 
Spark transformations include map, filter and join. To materialize an RDD, an
action must be performed, such as a reduce or a collect.

All actions and wide dependencies require a shuffle -- Spark's most costly
operation. Every shuffle begins with each map task saving its data to local
files to be fetched by the shuffle operation. The shuffle operation then 
redistributes the data across the partitions as requested. A shuffle marks a 
stage boundary in Spark. In other words, a shuffle will not begin until all 
dependent map tasks \todo{shouldn't it be 'actions'?} have completed.

Spark will spill RDD data to disk whenever the complete dataset does not fit in
memory. Moreover, as Spark transformations generate new RDDs and numerous 
transformations may occur in a single application, Spark implements a LRU eviction
policy. If an evicted RDD needs to be reused, Spark will recompute it using the
lineage data collected. It therefore becomes of great importance to persist to
memory and/or 
disk RDDs which will be reused throughout the application, particularly if the
recomputation is costly. 
%mention results in thunder paper

Data locality in Spark is achieved through the  scheduling of task to partitions which have the
data loaded in memory. If the data is instead stored on HDFS\todo{expand acronym if not done in intro}, the scheduler will
assign it to one of the preferred locations specified by HDFS. Spark's scheduler
utilizes delay scheduling to optimize fairness and locality for all tasks.
%improve paragraph

Three different types of schedulers are compatible with Spark: 1) Spark 
Standalone, 2) YARN~\cite{vavilapalli2013apache} and 3) Mesos~\cite{hindman2011mesos}. The Spark Standalone
scheduler is the default scheduler. YARN was a scheduler designed for Hadoop~\cite{white2012hadoop}
clusters. It is prepackaged with Hadoop installations. In contrast to YARN, 
Mesos was designed to be used in multitenant cluster environments. In our experiments,
we focus on the Spark Standalone cluster. \todo{Maybe defer this to experiments}

Executing Spark applications on HPC systems with Spark-unaware schedulers may
be inefficient. The amount of resources requested by Spark may impede Spark-cluster
scheduling time. Using pilot-scheduling strategies to add nodes to the Spark cluster
as they are allocated by the underlying schedulers may speedup allocation and 
overall processing time. This, however, is not studied in the current paper. \todo{Explain why}

@@

Nipype is a popular Python neuroimaging processing engine. It aims at being a 
solution for easily creating reproducible neuroimaging workflows. Although Nipype
does not employ any Big Data processing strategies, it provides access to numerous
plugins to schedulers found in most clusters readily available to researchers, 
such as SGE, PBS, SLURM and HTCondor\todo{expand acronyms}. It also includes its own scheduler, MultiProc,
for parallel processing on single nodes. Nipype also includes many built-in 
interfaces to commonly used neuroimaging tools to be incorporated within the 
workflows. Nipype's ability to easily parallelize workflows in researcher-available 
cluster setups, capture detailed provenance information necessary for reproducibility,
and allow users to easily integrate existing neuroimaging tools, make it preferable\todo{not the right word and sentence is difficult to parse.}
to existing Big Data solutions, which would necessitate modification to achieve this.

Jobs, or Interfaces, in Nipype, are encapsulated by a Node. A Node dictates that
the job will execute on a single input. However, the MapNode, 
a child variant of the Node, can execute on multiple inputs. All tasks in
Nipype execute in their own uniquely named subdirectory which facilitates the provenance 
tracking of inputs and outputs. This also enables checkpointing of the workflow.
In the case of Node failure or application modification, only the nodes who have
been modified (verified by hash) are re-executed.

In order for Nipype to operate as intended, a filesystem shared by all the nodes 
is required. However, it is still possible to save to a non-shared local filesystem, 
but it may come at the expense of fault-tolerance \todo{explain why}. Moreover, the user will need to
ensure that the files are appropriately directed to the nodes that require them as
there is no guarantee of data locality in Nipype.
